Distributed Topology Discovery in Self-Assembled Nano Networks

### 2014-02-14 PAPER/introduction.tex
Fixes

Exploring long term alternatives to the CMOS technology is gaining
more and more revelance as the scaling trend of such devices keeps
introducing new challenges. Power density, defect tolerance, testing
costs and wire delays are only a few of the many critical aspects
involved~\cite{itrs12}. While software parallelism and multicore
approaches~\cite{horowitz2004, powell2009} are partially mitigating the impact of such
constraints on performances, it is likely that the growing computing demand will
eventually need even more radical architectural modifications and new
paradigms in order to address the Computer Design challenges of the next
decades.

In the last years, Self-assembled nanoscale architectures~\cite{yan2003}
emerged as a promising technology due their tera/peta scale of
integration, defect tolerance and huge potential computing
capabilities. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{patwardhan2004, patwardhan2006_1, pistol2009}.
%, patwardhan2006
The main idea behind this approach is to exploit the physical regularity and
stability of DNA structures in order to create a scaffold onto which
nano-devices (e.g. nanowires and CNFETs~\cite{bachtold2001, tans1998, cui2001}) can be
attached. This can be achieved by designing appropriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold.
A detailed description of the chemical properties involved is far
beyond the scope of this paper (see also~\cite{braun1998, seeman1999}), so we will focus on
the three main challenges that this new fabrication process introduces
in Computer Design: \emph{(i) limited node complexity}, \emph{(ii) large scale
randomness} and \emph{(iii) high defect rates}.

@@

\begin{figure*}
\centering
\begin{tabular}{cc}
    \includegraphics[width=0.40\textwidth]{pictures/dna2b.eps} &
    \includegraphics[width=0.40\textwidth]{pictures/dna1_complex2.eps} \\
 (a) & (b)
 \end{tabular}
  \caption{DNA tag matching and Resulting Irregualr Topology}
  \label{fig:nana}
\end{figure*}
These aspects of a DNA-self assembly process lead to some important
implications to be addressed when approaching to the design of
computer architecture: a computational model should be based on a
distributed network of small computing and storage nodes, randomly
placed and interconnected (see Figure~\ref{fig:nana}).
Proof-of-concept of such kind of computational model can be found for
example in~\cite{patwardhan2006_1}.  Since no regularity can be
assumed in such network, a topology agnostic strategy that avoids
deadlock should be adopted in order to route packets (e.g.
instructions and data operands) among nodes.  It is important to
underline that, altougth we can assume an architecture similar to the
one depicted in Figure~\ref{fig:nana} as background, the topology discovery
problem faced in this work is more general and not strictly dependent on any of the
underlying computing architecture.

In this work we introduce \disr{}, a Distributed Segment-based approach
to deadlock freedom in large scale DNA self assembled networks. Our
contribution aims to achieve the classical properties of a
segment-based approach~\cite{mejia_ipdps06} without requiring any
topology graph, external defect map or centralized algorithm
execution.  So the \disr{} approach is not intended to discover the
``optimal'' segment choice (ideally reachable with the knowledge of
the topology graph) but just to demonstrate a concrete model that can
fit into such complex, irregular and large sized networks.

@@

\section{Related Works and Contribution}
\label{sec:related_works}
Several works~\cite{sancho2002, skeie2002, skeie2004, koibuchi2003} have been presented addressing the problem of topology
agnostic routing. While they show interesting
performances, the requirement for virtual channels resources 
to avoid deadlock makes them not suitable for the specific DNA self
assembled networks we are assuming as background.
Other topology independent algorithms~\cite{schroeder1991, koibuchi2001, cherkasova1996} do not use virtual channels to
achieve deadlock freedom, but turn prohibitions. 
In particular, authors of~\cite{Patwardhan05evaluatingthe} exploits the creation of a spanning tree of the
topology, then placing bidirectional restrictions by avoiding a packet
to traverse the same link in both up and down directions.
While the hierarchical nature of this approach can lead to uneven traffic
distribution, with many packets traversing upper links (near to the
root), this is quite acceptable in classical wide area networks
topologies with a limited number of nodes. Other approaches such as
FX~\cite{sancho2000} mitigate this
issue, but the set of turn restrictions is still prefixed,
strictly depending on the particular tree root selected. 
Other solutions try to approach the issue of irregular
topologies by limiting the number~\cite{dally1994, duato1997, gomez2004, koibuchi2008} or the
location~\cite{zhang2008, sui2000, flich2008, liu2011} of missing links. This
restriction is clearly unacceptable given the hight-defect rates of
DNA self-assembled networks scenario. For the same reason, we also
avoid considering solutions based on hardware-redundancy to
dynamically recover defects as in~\cite{constantinides2006, kohler2010, kim2006, park2006, ebrahimi2013}.

@@

The \disr{} approach presented in this work is a first attempt of
reaching the same goals of a segment based turn prohibition  without
requiring any topology graph or centralized execution.  In particular
\disr{} aims to obtain a segment network partitioning using a
completely different mechanism, based on a distributed exchange of
small setup packets. The following main features distinguish \disr{}
from classic approaches to segment partitioning: 
\begin{itemize}
\item No centralized entity is globally responsible and/or aware of
what is going on, i.e. the status of the \disr{} execution is
collectively distributed among the nodes
\item  No defect map and/or topology graph is used as input,
thus, the topology has to be discovered \emph{while} segments are
created
\item At the end of the execution, no segment list is created or
stored anywhere: each node is only aware of belonging to some segment,
ignoring the presence of other nodes in the same segment and even the
presence of other segments in the network
\end{itemize}

>>>>

Exploring long term alternatives to the CMOS technology is gaining
more and more revelance as the scaling trend of such devices keeps
introducing new challenges: power density, defect tolerance, testing
costs and wire delays are only a few of the many critical aspects
involved~\cite{itrs13}. 
While software parallelism and multicore
approaches~\cite{horowitz2004, powell2009} are partially mitigating the impact of such
constraints on performances, it is likely that the growing computing demand will
eventually need even more radical architectural modifications and new
paradigms in order to address the Computer Design challenges of the next
decades.

In the last years, Self-assembled nanoscale architectures~\cite{yan2003}
emerged as a promising technology due their tera/peta scale of
integration, defect tolerance and huge potential computing
capabilities. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{patwardhan2004, patwardhan2006_1, pistol2009}.
%, patwardhan2006
The main idea behind this approach is to exploit the physical regularity and
stability of DNA structures in order to create a scaffold onto which
nano-devices (e.g. nanowires and CNFETs~\cite{bachtold2001, tans1998, cui2001}) can be
attached. This can be achieved by designing appropriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold (see Figure~\ref{fig:nana}-a).
A detailed description of the chemical properties involved is far
beyond the scope of this paper (see also~\cite{braun1998, seeman1999}), so we will focus on
the three main challenges that this new fabrication process introduces
in Computer Design: \emph{(i) limited node complexity}, \emph{(ii) large scale
randomness} and \emph{(iii) high defect rates}.

@@

\begin{figure*}
\centering
\begin{tabular}{cc}
    \includegraphics[width=0.40\textwidth]{pictures/dna2b.eps} &
    \includegraphics[width=0.40\textwidth]{pictures/dna1_complex2.eps} \\
 (a) & (b)
 \end{tabular}
  \caption{(a) DNA tag matching and (b) resulting trregualr topology}
  \label{fig:nana}
\end{figure*}
These aspects of a DNA-self assembly process lead to some important
implications to be addressed when approaching to the design of
a computer architecture: computational model should be based on a
distributed network of small processing and storage nodes, randomly
placed and interconnected (see Figure~\ref{fig:nana}-b).
Proof-of-concept of such architectures can be found for
example in~\cite{patwardhan2006_1}, where instruction and data
operands are move around the network in order to be processed.
Indepentedly from the particular policy chosen for routing packets,
such networks must be able to guarantee deadlock freedom without any
prior assumption on topology.

In this work we introduce \disr{}, a Distributed Segment-based approach
to deadlock freedom in large scale DNA self assembled networks. Our
contribution aims to achieve the classical properties of a
segment-based approach~\cite{mejia_ipdps06} without requiring any
topology graph, external defect map or centralized algorithm
execution.  So the \disr{} approach is not intended to discover the
``optimal'' segment choice (ideally reachable with the knowledge of
the topology graph) but just to demonstrate a concrete model that can
fit into such complex, irregular and large sized networks.
It is important to
underline that, altougth we can assume an architecture similar to the
one depicted in Figure~\ref{fig:nana} as background, the topology discovery
problem faced in this work is more general and not strictly dependent on any of the
underlying computing architecture.

@@

\section{Background and Contribution}
\label{sec:related_works}
Several worksaddressed the problem of topology agnostic deadlock
freedom introducing virtual channels~\cite{sancho2002, skeie2002,
skeie2004, koibuchi2003}, but while they show interesting
performances, the low resources deadlock makes them not suitable for
the specific DNA self assembled networks we are assuming as
background.
Other topology independent algorithms~\cite{schroeder1991,
koibuchi2001, cherkasova1996} do not require any virtual channels but
exploit the concept of ``turn prohibition''.
In particular, authors of~\cite{Patwardhan05evaluatingthe} exploits the creation of a spanning tree of the
topology, placing bidirectional restrictions by avoiding a packet
to traverse the same link in both up and down directions.
The hierarchical nature of this approach can lead to uneven traffic
distribution, with many packets traversing upper links (near to the
root), but this is quite acceptable in classical wide area networks
topologies with a limited number of nodes. Other approaches such as
FX~\cite{sancho2000} mitigate this
issue, but the set of turn restrictions is still prefixed,
strictly depending on the particular tree root selected. 
Other solutions try to approach the issue of irregular
topologies by limiting the number~\cite{duato1997, gomez2004, koibuchi2008} or the
location~\cite{zhang2008, flich2008, liu2011} of missing links. This
restriction is clearly unacceptable given the hight-defect rates of
DNA self-assembled networks scenario. For the same reason, we also
avoid considering solutions based on hardware-redundancy to
dynamically recover defects as in~\cite{constantinides2006, kohler2010,  park2006, ebrahimi2013}.

@@

The \disr{} approach presented in this work is a first attempt of
achieving the same goals of a segment based turn prohibition  without
requiring any topology graph or centralized execution. In particular
\disr{} implements segment discovery and turn prohibition using a
completely different mechanism, based on a distributed exchange of
small setup packets requiring on average $15\%$ of the available
resource budget. The following main features distinguish \disr{}
from classic approaches to segment partitioning: 
\begin{itemize}
\item No centralized entity is globally responsible and/or aware of
what is going on, i.e. the status of the \disr{} execution is
collectively distributed among the nodes
\item  No defect map and/or topology graph is used as input,
thus, the topology has to be discovered \emph{while} segments are
created
\item At the end of the execution, no segment list is created or
stored anywhere: each node is only aware of belonging to some segment,
ignoring the presence of other nodes in the same segment and even the
presence of other segments in the network
\end{itemize}

### 2014-02-14 PAPER/paper.tex
Fixes

\newcommand{\etal}{\emph{et al.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\etc}{\emph{etc.}}
\renewcommand{\baselinestretch}{0.98}\normalsize
%------------------------------------------------------------------------------

@@

\begin{abstract}
In this paper we present DiSR, a first effort for a distributed
segment-based approach to routing and defect mapping in a nano-scale,
topology agnostic scenario based on DNA self-assembly. The main aim is
exploiting the already-proven qualities of segment-based routing
without neither requiring a topology graph as input, nor needing a
centralized algorithm to configure network paths.  After introducing
the conceptual elements and the execution model of DiSR, we show how
the opensource Nanoxim platform has been used to evaluate the proposed
approach in the process of discovering irregular network topology
while establishing network segments.  Results show how DiSR still
preserves some important properties (coverage, defect tolerance,
scalability)  while avoiding centralized tree-based broadcasting and
resource hungry solutions such as virtual channels and hardware
redundancy. Finally, we analyzed a first, not yet optimised gate-level
hardware implementation of the required control logic and storage for
DiSR, demonstrating a relatively acceptable impact ranging from 10 to
about 20\% of the budget of transistors available for each node.
\end{abstract}

@@

\section{Conclusions}
In this work we presented DiSR, an initial attempt of porting a segment based
approach to distributed self-assembled nanoscale scenario. We
demonstrated how DiSR can accomplish the aim of finding a segment
coverage of the network without requiring any centralized approach
that would request a graph topology as input. A draft hardware
implementation has been presented to evaluate the impact on the limited
node size typical of the assumed scenario. Future works will focus on
developing self tuning of DiSR to match a high coverage levels and the
detailed low level hardware implementation on
DNA grid using the appropriate nano devices models.

>>>>

\newcommand{\etal}{\emph{et al.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\etc}{\emph{etc.}}
\newcommand{\disr}{{\sf DiSR}}
%\renewcommand{\baselinestretch}{0.97}\normalsize
%------------------------------------------------------------------------------

@@

\begin{abstract}
In this paper we present \disr{}, a distributed approach to topology
discovery and defect mapping in a self-assembled nano
networks-on-chip. The main aim is achieving the already-proven
properties of segment-based deadlock freedom without neither requiring
a topology graph as input, nor needing a centralized algorithm to
configure network paths.  After introducing the conceptual elements
and the execution model of \disr{}, we show how the opensource Nanoxim
platform has been used to evaluate the proposed approach in the
process of discovering irregular network topology while establishing
network segments. Comparison against a tree-based approach show how
\disr{} still preserves some important properties (coverage, defect
tolerance, scalability) while avoiding resource hungry solutions such
as virtual channels and hardware redundancy. Finally, we propose a
gate-level hardware implementation of the required control logic and
storage for \disr{}, demonstrating a relatively acceptable impact
ranging from 10 to about 20\% of the budget of transistors available
for each node.
\end{abstract}

@@

\section{Conclusions}
In this work we presented \disr{}, an initial attempt of achieving a segment based
topology discovery into distributed self-assembled nanoscale scenario. We
demonstrated how \disr{} can accomplish the aim of finding a segment
coverage of the network without requiring any centralized approach
that would request a graph topology as input. A draft hardware
implementation has been presented to evaluate the impact on the limited
node size typical of the assumed scenario. From an high-level
perspective future works will focus on
investigating \disr{} resulting network in order to support the
execution of massively parallel applications, while keeping into
developing a detailed low level hardware implementation on
DNA grid using the appropriate nano devices models.
